Move Schedules: Fast persistence computations in coarse dynamic settings\thanks{\textbf{Funding:} This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758.}

### 2021-10-26 move_schedules.tex
Updates from Overleaf

% Authors: full names plus addresses.
\author{
	Matt Piekenbrock\thanks{Department of Computational Mathematics, Science \& Engineering, Michigan State University.}
	\and Jose A. Perea\thanks{
	Department of Computational Mathematics, Science \& Engineering, 
	Department of Mathematics, Michigan State University.}
}

@@

\section{Introduction} 
\subsection{Overview}\label{sec:overview} 
Given a triangulable topological space equipped with a sufficiently tame continuous function, persistent homology captures the changes in topology across the sublevel sets of the space, and encodes them in a persistence diagram. The stability of persistence contends that small changes in the underlying function produce small changes in the corresponding diagrams:  if the function changes continuously, so too will the points on the persistence diagram~\cite{cohen2007stability, cohen2006vines}. 
This motivates the study and application of persistence to time-varying or dynamic settings, like that of dynamic metric spaces~\cite{kim2020spatiotemporal}. 
As persistence-related computations tend to exhibit high algorithmic complexity---essentially cubic\footnote{For finite fields, it is known that the persistence computation reduces to the PLU factorization problem, which takes $O(m^\omega)$ where $\omega \approx 2.373$ is the matrix multiplication constant.} in the size of the underlying filtration~\cite{morozov2005persistence}---their adoption to dynamic settings poses a challenging computational problem.
With the current state of the art tools, there is no recourse to the problem of computing persistence from a time-varying filtration containing millions of simplices at thousands of snapshots in time.
Efficiently  updating the persistence of a time-varying filtration has far-reaching consequences: methods that vectorize persistence diagrams for machine learning purposes, like adaptive template functions~\cite{polanco2019adaptive} and persistence images~\cite{adams2017persistence}, or summary statistics like $\alpha$-smoothed Betti curves~\cite{ulmer2019topological}, all immediately become computationally viable tools in dynamic  settings. 
%This analogy provides a strong interpretation at gaining insight on the subtle topological changes that may occur in studying a continuous process.

@@

Despite its theoretical efficiency, vineyards is often not the method of choice in practical settings. 
Indeed, while there is an increasingly rich ecosystem of software packages offering variations of the standard reduction algorithm (e.g. Ripser, PHAT, Dionysus, etc... see~\cite{otter2017roadmap} for an overview), implementations of the vineyards algorithm are relatively uncommon.\footnote{Dionysus 1 has an implementation of vineyards, but the algorithm was never ported to the 2nd version.} 
The reason for this disparity is perhaps explained by Lesnick and Wright~\cite{lesnick2015interactive}: ``While an update to an $RU$ decomposition involving few transpositions is very fast in practice... many transpositions can be quite slow... it is sometimes much faster to simply recompute the $RU$-decomposition from scratch using the standard persistence algorithm.'' 
Whether performing the reductions naively or updating the decomposition dynamically using vineyards, they observe that this is the most computationally demanding aspect of  RIVET \cite{rivet}.

This work seeks to further understand and remedy this discrepancy: building on the work presented in~\cite{busaryev2010tracking}, we introduce a coarser approach to the vineyards algorithm.
While vineyards is  designed  for handling   \emph{continuous} time families of diagrams, it is not necessarily efficient when the time parameter is coarsely discretized.
Our methodology is based on the observation that practitioners often don't need (or want!) all $d$ persistence diagrams generated by a continuous 1-parameter of filtrations; usually just $n << d$ of them are sufficient.   
% in many time varying settings, a valid decomposition is often only sought after at a relatively sparse number of time points relative to the size of the (finite) set needed to generated by the family. 
By exploiting the ``donor'' concept introduced in in~\cite{busaryev2010tracking}, we are able to make a tradeoff in the number of times the decomposition is restored to a valid state, dramatically reducing the total number of column operations needed to apply an arbitrary permutation to the filtration. This trade off, paired with a fast greedy heuristic explained in section~\ref{sec:proxy_objective}, yields an algorithm that can update a $R = DV$ decomposition more efficiently than vineyards in coarse contexts, making dynamic persistence more computationally tractable for a wider class of use-cases.

@@

 \subsection{A Motivating Example and Contributions}\label{sec:motivation} 
As a motivating example, we describe a simple experiment illustrating why the vineyards algorithm does not always yield an efficient method for time-varying applications. 
Consider a series of grayscale images (i.e., a video) depicting a fixed-width annulus expanding about the center of a $9 \times 9$ grid, where the scale of the annulus is parameterized by a single parameter $\tau$. 
Each image contains the same number of pixels whose intensities---these are determined by $\tau$---vary with time. For each image, we build a filtered simplicial complex using the \emph{Freudenthal} triangulation of the plane, with the order of inclusions obtained by lower stars.  
 \begin{figure*}[t]
 	\centering
 	\includegraphics[width=0.36\textwidth, height = 1.45in]{circle_vineyards.png}
 	\includegraphics[width=0.63\textwidth, height = 1.45in]{circle_complex.png}
 	\caption{On the right, the grayscale image data set and corresponding lower star filtrations. On the left, the cumulative number of column operations required to compute persistence for this time-varying filtration. The   Betti numbers for each of the shown  steps in the filtration, from left to right, are: $(\beta_0,\beta_1, \beta_2)  = (1,0,0), \; (1,1,0),  \; (1,1,0), \; (1,1,0), \; (4,0,0)$. Observe that computing the diagrams independently at 10 evenly spaced time points (the green line) both captures these major topological changes and is the most computationally efficient approach shown.}
 	\label{fig:vineyards}
 \end{figure*}
Snapshots of the image data and its corresponding time-varying filtration is depicted in Figure~\ref{fig:vineyards} (top right and bottom right, respectively), which we use as a baseline for comparing the standard reduction algorithm $\texttt{pHcol}$ (Algorithm \ref{alg:reduce}) with vineyards. Note that the size of the filtration is fixed--no simplices are ever added or removed over time. 
There are two events which dramatically alter the persistence diagrams extracted from these filtrations: the first occurs when the central connected component splits to form a cycle, and the second occurs when the annulus approaches the outer dimensions of the grid, splitting into four components. As a consequence, in this example, only a few persistence diagrams are needed to capture the major topological changes over time.

  Consider the situation where one is concerned with capturing these dramatic changes in homology.
  On the left side of Figure~\ref{fig:vineyards}, we compare the cumulative cost (in total number of column operations) of various approaches.
  Since it is unknown a priori at which times  the persistent homology will change, one solution is to discretize the time domain at $n$ evenly spaced points and compute the persistence diagrams at each one of them  independently. An alternative approach is to construct a homotopy between two contiguous time-points, and then decompose this homotopy into adjacent transpositions--the vineyards approach. The former is often used in practice, but it is only an approximation; the latter is guaranteed to capture all homological changes in persistence that occur in simulating the 1-parameter family. 
  The graph in Figure~\ref{fig:vineyards} depicts two different size approximations (purple, $n=100$ and green, $n=10$), wherein the reduction algorithm (\texttt{pHcol}) was applied at $n$ time points, and two exact strategies (black and blue), which compute all $\approx 7 \times 10^{4}$ diagrams given by the homotopy. 
  As one can see from the figure, if one needs to simulate the homotopy exactly, then the vineyards approach is indeed far more efficient than naively applying the reduction algorithm independently at all time points.
  However, when the discretization of the time domain is coarse enough, the naive approach actually performs less column operations than the vineyards strategy, while still capturing the main events.

  The existence of a time discretization that is more efficient to compute than continually updating the decomposition, indicates that the vineyards algorithm must incur some amount of overhead (in terms of column operations) to perform transpositions, even if the persistence diagrams remain unchanged. 
  Even in situations where one is updating a decomposition to a filtration that is  relatively ``close'', e.g. any two adjacent filtrations in the bottom right of Figure~\ref{fig:vineyards}, it is still more efficient to apply \texttt{pHcol} directly than  to iteratively update the decomposition (see the case where $n = 10$). 
 Moreover, any optimization to the reduction algorithm (e.g.,  clearing~\cite{chen2011persistent}) would only increase this disparity.
 See~\cite{topaz2015topological, xian2020capturing, lesnick2015interactive, kim2020spatiotemporal} 
 for relevant scenarios and applications.

\textbf{Our approach and contributions are as follows:} First, we leverage the \emph{moves} framework of Busaryev et. al.~\cite{busaryev2010tracking} to include  coarser  operations---in terms of valid intermediate decomposition states, as compared to vineyards---for the dynamic setting. 
We give a tight lower bound on the number of moves needed to perform an arbitrary permutation to the $R = D V$ decomposition, and give a proof of optimality by a reduction to the permutation edit distance problem.
We also give worst-case bounds in expectation as well as efficient algorithms for achieving these bounds---both of which are derived from a reduction to the Longest Increasing Subsequence (LIS) problem. 
This reduction yields an efficient algorithm for generating sequences of moves $( s_1, s_2, \dots, s_d )$ of minimal size $d$, which we call \emph{schedules}. 
However, not all valid schedules incur the same permutation costs 
(i.e., number of column operations).
We investigate the feasibility of choosing optimal cost schedules, and show that 
greedy-type approaches can lead to arbitrarily bad behavior. 
In light of these results, we give an alternative proxy-objective for cost minimization, provide bounds justifying its relevance to the original problem, and give an efficient $O(d \log d)$ algorithm for approximately solving this proxy minimization. 
A performance comparison with other reduction-based persistent homology computations is given; 
our move schedules are demonstrated to be an order of magnitude more efficient than existing approaches in applications such as flock simulation and 2D persistence computations. 
%We conclude with a discussion of situations where our strategy is most applicable and future directions for research.

@@

\subsection{Outline} The remainder of the paper is organized as follows: we review and establish the notations we will use to describe simplicial complexes, persistent homology, and dynamic persistence in Section \ref{sec:background}. 
We also cover the reduction algorithm (designated here as \texttt{pHcol}), the vineyards algorithm, and the set of \emph{move}-related algorithms introduced in~\cite{busaryev2010tracking}, which serves as the starting point of this work. 
In Section \ref{sec:move_schedules} we introduce our move schedules  and provide efficient algorithms to make this alternative viable. 
In Section \ref{sec:results} we demonstrate a few applications of the proposed method. 
In particular, to the computation of crocker stacks from flock simulations, and to 2-dimensional persistence on the Mumford data set of image patches from natural images. 
In Section \ref{sec:conclusion} we conclude the paper by discussing other possible applications and future work. 
We also include an appendix which introduces certain topics in more detail, such as the vineyards algorithm itself, for readers who may be unfamiliar but would like a more in-depth understanding.

@@

The $p$-boundary $\partial_p(\sigma)$  of a $p$-simplex $\sigma\in K$ is the alternating sum of its oriented co-dimension 1 faces; the $p$-boundary of a $p$-chain is defined 

@@

If $\{K_i\}_{i\in [m]}$ is a filtration, then the inclusion maps  $K_i\subset K_j$  for $i < j$, induce linear transformations $f_p^{i,j}: H_p(K_i) \to H_p(K_j)$ at the level of homology:

@@

The $p$-th persistent homology groups are the images of these transformations: $H_{p}^{i,j} = \mathrm{Im}\,f_p^{i,j}$. 

@@

Given a triangulable topological space $\mathbb{X}$ equipped with a
  function $f: \mathbb{X} \to \mathbb{R}$, 
we write $\mathbb{X}_a = f^{-1}(-\infty, a]$ to denote the sublevel sets of $\mathbb{X}$ defined by the value $a$. 
A \emph{homological critical value} of $f$ is any value $a \in \mathbb{R}$ such that the homology of the sublevel sets of $f$ changes at $a$, i.e.  if for some $p$ the inclusion-induced homomorphism  $H_p(\mathbb{X}_{a - \epsilon}) \to H_p(\mathbb{X}_{a+\epsilon})$ is not an isomorphism for  any $\epsilon >0$. If there are only finitely many of these homological critical values, then $f$ is said to be \emph{tame}.

Consider a homotopy $F(x,\tau) : \mathbb{X} \times [0,1] \to \mathbb{R}$ and denote its ``snapshot'' at a given time-point $\tau$ by $f_\tau(x) = F(x,\tau)$.
The snapshot $f_0$ denotes the initial function at time $\tau = 0$ and $f_1$ denotes the function at the last time step. 
As $\tau$ varies in $[0,1]$, the points in $\mathrm{dgm}_p(f_\tau)$ trace  curves in $\mathbb{R}^3$, called \emph{vines} and which together form a \emph{vineyard}. 
The stability of persistence implies that these curves will be continuous if the homotopy is continuous.
The vineyard analogy acts as a guidepost for practitioners seeking to understand how subtle changes occurring to the topological structure over time.
A more detailed discussion of the vineyards approach is provided in section~\ref{sec:vineyards}.
%Let $\mathbb{M}$ be a smooth $d$-manifold and $f : \mathbb{M} \to \mathbb{R}$ a smooth function. 
%A critical point $x$ is called \emph{non-degenerate} if the Hessian matrix of second partial derivatives is non-singular, otherwise if the Hessian is zero $x$ is said to be \emph{degenerate}. A \emph{Morse function} is a smooth function that has only non-degenerate critical points with distinct critical values. 
%and two Morse functions $f_0, f_1: \mathbb{M} \to \mathbb{R}$, and further suppose we equip $f_0, f_1$ with with a smooth homotopy $F : \mathbb{M}: [0,1] \to \mathbb{R}$ such that $F(x, 0) = f_0(x)$ and $F(x, 1) = f_1(x)$ for $x \in \mathbb{M}$. We use $f_\tau$ to denote the function at $F(x, \tau)$, where $\tau \in [0, 1]$. 
\subsection{The Reduction Algorithm}\label{sec:reduction}
In this section we briefly recount the original reduction algorithm introduced in~\cite{zomorodian2005computing}, also sometimes called the \emph{standard} algorithm or more explicitly \texttt{pHcol}~\cite{de2011dualities}. 
The pseudocode is outlined in Algorithm~\ref{alg:reduce}. Without optimizations, like   clearing   or       implicit matrix reduction, the standard  algorithm is very inefficient. Nonetheless, it serves as the foundation of most implementations that compute persistent homology, and its invariants are necessary before introducing both vineyards in section~\ref{sec:vineyards} and our move schedules  in section~\ref{sec:move_schedules}.

@@

Using $\mathbb{Z}_2$ coefficients and homology in dimension 1 to simplify the presentation, the steps look at follows:

@@

\noindent The column $c$ in $R$ indicates that dimension $1$ homology is born, which in this filtration is never killed since the dimension of the complex is $2$. Similarly, the columns at $u, v, w$ in $R_0$ (not shown) are all zero, indicating three $0$-dimensional homology classes  are born, which are then killed by the pivot entries in columns $a$ and $b$ in $R_1$.

@@

Its clear from Algorithm~\ref{alg:reduce} that a loose upper bound for this reduction is $O(m^3)$, where $m$ is the number of simplices of the filtration; it turns out that this bound is in fact tight, see~\cite{morozov2005persistence} for more details. 

@@

Notice that homological critical events can only occur when the filtration order changes, though the converse isn't true.
Thus, detecting these all such events reduces to computing matrix decompositions at a finite set of time points interleaved between changes in the filtration order. 
At the finest scale, these changes manifest as adjacent transpositions. 
As a result, a fixed set of rules for maintaining a valid $R = D V$ decomposition under the transposition of adjacent simplices in the filtration is enough to simulate persistence dynamically. 

@@

 The time complexity of vineyards is determined entirely by the complexity of performing a single adjacent transposition, and its complexity is often stated in this way. Informally, since column operations are the largest complexity operations needed and each column can have potentially $m$ entries, the complexity of vineyards is $O(m)$ per transposition. 
 However, there are a number of subtleties involved in this statement. 
 The first subtlety is that both the $V$ and $R$ matrices tend to be quite sparse, and thus column and row operations rarely require $m$ operations---in fact, as a rule of thumb, most transpositions require no column operations~\cite{edelsbrunner2000topological}. A second subtlety, due to the necessity of sparse matrix data structures, is that querying the non-zero status of any particular entry typically takes $O(\log m)$ time. This hidden cost may seem minor, however, the quadratic scaling of the number of transpositions vineyards typically requires makes this cost non-trivial.   
In fact, achieving $O(m)$ complexity per transposition requires a special sparse matrix representation that allows swapping any two rows and columns in $O(1)$---see the appendix for more details and discussion. Finally, note the constant factor associated with the stated complexity: inspection of the individual cases of the algorihtm from~\cite{cohen2006vines} shows that there are at most two $O(m)$ operations, on both $R$ and $V$, that may be needed for any single transposition.

@@

Though they are similar, move operations confer a few computational advantages compared to vineyards. Informally, a move operation has two convenient properties:
\\	
 \begin{enumerate}[leftmargin=2\parindent, align=left, labelsep=-5pt, topsep=0pt,itemsep=-0.25ex,parsep=1.2ex]
 	\item[M1:] Querying the non-zero status of entries in either $R$ or $V$ occurs once per move.
 	\item[M2:] $R = D V$ is not maintained during the movement of $\sigma_i \mapsto \sigma_j$.
 \end{enumerate} 
 \vspace*{1em}
First, consider property M1. Prior to applying any permutation $P$ to the decomposition, it is necessary to remove non-zero entries in $V$ which render $P^TVP$ non-upper triangular, to maintain invariant I2. 
If one uses the (augmented) sparse matrix representation described in the original vineyards publication~\cite{cohen2006vines}, checking whether these entries are non-zero takes $O(\log m)$ per column. Since moving a simplex from $i$ to $j$ using vineyards performs $\lvert i - j \rvert - 1$ transpositions, this query operation is required at most $O(\lvert i - j \rvert)$ times, which is excessive. If a row-oriented sparse matrix implementation is used, a move operation can perform these non-zero entry checks in just one $O(m)$ pass, prior to performing any column operations.

>>>>

% Authors: full names plus addresses.
\author{
	Matt Piekenbrock\thanks{Khoury College of Computer Sciences, Northeastern University.}
	\and Jose A. Perea\thanks{
	Department of Mathematics and Khoury College of Computer Sciences, Northeastern University.}
}

@@

\section{Introduction} 
\subsection{Overview}\label{sec:overview} 
Given a triangulable topological space equipped with a  tame continuous function, persistent homology captures the changes in topology across the sublevel sets of the space, and encodes them in a persistence diagram. The stability of persistence contends that   if the function changes continuously, so too will the points on the persistence diagram~\cite{cohen2007stability, cohen2006vines}. 
This motivates the  application of persistence to time-varying  settings, like that of dynamic metric spaces~\cite{kim2020spatiotemporal}. 
As persistence-related computations tend to exhibit high algorithmic complexity---essentially cubic\footnote{For finite fields, it is known that the persistence computation reduces to the PLU factorization problem, which takes $O(m^\omega)$ where $\omega \approx 2.373$ is the matrix multiplication constant.} in the size of the underlying filtration~\cite{morozov2005persistence}---their adoption to dynamic settings poses a challenging computational problem.
With the current state of the art tools, there is no recourse to the problem of computing persistence from a time-varying filtration containing millions of simplices at thousands of snapshots in time.
Efficiently  updating the persistence of a time-varying filtration has far-reaching consequences: methods that vectorize persistence diagrams for machine learning purposes, like adaptive template functions~\cite{polanco2019adaptive} and persistence images~\cite{adams2017persistence}, or summary statistics like $\alpha$-smoothed Betti curves~\cite{ulmer2019topological}, all immediately become computationally viable tools in dynamic  settings. 
%This analogy provides a strong interpretation at gaining insight on the subtle topological changes that may occur in studying a continuous process.

@@

Despite its theoretical efficiency, vineyards is often not the method of choice in practical settings. 
Indeed, while there is an increasingly rich ecosystem of software packages offering variations of the standard reduction algorithm (e.g. Ripser, PHAT, Dionysus, etc. See~\cite{otter2017roadmap} for an overview), implementations of the vineyards algorithm are relatively uncommon.\footnote{Dionysus 1 has an implementation of vineyards, but the algorithm was never ported to the 2nd version.} 
The reason for this disparity is perhaps explained by Lesnick and Wright~\cite{lesnick2015interactive}: ``While an update to an $RU$ decomposition involving few transpositions is very fast in practice... many transpositions can be quite slow... it is sometimes much faster to simply recompute the $RU$-decomposition from scratch using the standard persistence algorithm.'' 
Whether performing the reductions naively or updating the decomposition dynamically using vineyards, they observe that this is the most computationally demanding aspect of  RIVET \cite{rivet}.

This work seeks to further understand and remedy this discrepancy: building on the work presented in~\cite{busaryev2010tracking}, we introduce a coarser approach to the vineyards algorithm.
While vineyards is  designed  for handling   \emph{continuous} time families of diagrams, it is not necessarily efficient when the time parameter is coarsely discretized.
Our methodology is based on the observation that practitioners often don't need (or want!) all $d$ persistence diagrams generated by a continuous 1-parameter of filtrations; usually just $n << d$ of them  suffice.   
% in many time varying settings, a valid decomposition is often only sought after at a relatively sparse number of time points relative to the size of the (finite) set needed to generated by the family. 
By exploiting the ``donor'' concept introduced in in~\cite{busaryev2010tracking}, we are able to make a tradeoff in the number of times the decomposition is restored to a valid state, dramatically reducing the total number of column operations needed to apply an arbitrary permutation to the filtration. This trade off, paired with a fast greedy heuristic explained in section~\ref{sec:proxy_objective}, yields an algorithm that can update a $R = DV$ decomposition more efficiently than vineyards in coarse contexts, making dynamic persistence more computationally tractable for a wider class of use-cases.

@@

 \subsection{A Motivating Example and Contributions}\label{sec:motivation} 
As a motivating example, we describe a simple experiment illustrating why the vineyards algorithm does not always yield an efficient strategy for time-varying applications. 
Consider a series of grayscale images (i.e., a video) depicting a fixed-width annulus expanding about the center of a $9 \times 9$ grid. 
Each image contains the same number of pixels, whose intensities vary with time. For each image, we build a filtered simplicial complex using the \emph{Freudenthal} triangulation of the plane, with the order of inclusions obtained by lower stars and pixel values.  
 \begin{figure*}[t]
 	\centering
 	\includegraphics[width=0.36\textwidth, height = 1.45in]{circle_vineyards.png}
 	\includegraphics[width=0.63\textwidth, height = 1.45in]{circle_complex.png}
 	\caption{On the right, the grayscale image data set and corresponding lower star filtrations. On the left, the cumulative number of column operations required to compute persistence for this time-varying filtration. The   Betti numbers for each of the shown  steps in the filtration, from left to right, are: $(\beta_0,\beta_1, \beta_2)  = (1,0,0), \; (1,1,0),  \; (1,1,0), \; (1,1,0), \; (4,0,0)$. Observe that computing the diagrams independently at 10 evenly spaced time points (the green line) both captures these major topological changes and is the most computationally efficient approach shown.}
 	\label{fig:vineyards}
 \end{figure*}
Snapshots of the image data and its corresponding time-varying filtration is depicted in Figure~\ref{fig:vineyards} (top right and bottom right, respectively), which we use as a baseline for comparing vineyards and the standard reduction algorithm $\texttt{pHcol}$ (Algorithm \ref{alg:reduce}).
Note that the size of the filtration is fixed---no simplices are ever added or removed over time. 
There are two events which dramatically alter the persistence diagrams extracted from these filtrations: the first occurs when the central connected component splits to form a cycle, and the second  when the annulus   splits into four components. As a consequence, in this example, only a few persistence diagrams are needed to capture the major topological changes over time.

  Consider the situation where one is concerned with capturing these dramatic changes in homology.
  On the left side of Figure~\ref{fig:vineyards}, we compare the cumulative cost (in total number of column operations) of various approaches.
  Since it is unknown a priori at which times  the persistent homology will change, one solution is to discretize the time domain at $n$ evenly spaced points and compute the persistence diagrams at each one of them  independently. 
  An alternative approach is to construct a (e.g., straight-line) homotopy between  contiguous time-points, and then decompose these homotopies into adjacent transpositions in the filtration order---the vineyards approach. The former is often used in practice, but it is only an approximation; the latter is guaranteed to capture all homological changes in persistence that occur in simulating the 1-parameter family. 
  The graph in Figure~\ref{fig:vineyards} depicts two different size approximations (purple, $n=100$ and green, $n=10$), wherein the reduction algorithm (\texttt{pHcol}) was applied at $n$ time points, and two exact strategies (black and blue), which compute all $\approx 7 \times 10^{4}$ diagrams given by the homotopy. 
  As shown in the figure, if one needs to simulate the homotopy exactly, then the vineyards approach is indeed far more efficient than naively applying the reduction algorithm independently at all time points.
  However, when the discretization of the time domain is coarse enough, the naive approach actually performs less column operations than the vineyards strategy, while still capturing the main events.

  The existence of a time discretization that is more efficient to compute than continually updating the decomposition, indicates that the vineyards algorithm must incur some overhead (in terms of column operations) to perform transpositions, even if the persistence diagrams remain unchanged. 
  Even in situations where one is updating a decomposition to a filtration that is  relatively ``close''---e.g. any two adjacent filtrations in the bottom right of Figure~\ref{fig:vineyards}---it is still more efficient to apply \texttt{pHcol} directly than  to iteratively update the decomposition (see the case where $n = 10$). 
 Moreover, any optimization to the reduction algorithm (e.g.,  clearing~\cite{chen2011persistent}) would only increase this disparity.
 See~\cite{topaz2015topological, xian2020capturing, lesnick2015interactive, kim2020spatiotemporal} 
 for relevant scenarios and applications.

\textbf{Our approach and contributions are as follows:} First, we leverage the \emph{moves} framework of Busaryev et. al.~\cite{busaryev2010tracking} to include  coarser  operations---in terms of the number of valid intermediate decomposition states, as compared to vineyards---for dynamic persistence. 
We give a tight lower bound on the number of moves needed to perform an arbitrary permutation to the $R = D V$ decomposition, and give a proof of optimality by a reduction to the permutation edit distance problem.
We also give worst-case bounds in expectation as well as efficient algorithms for achieving these bounds---both of which are derived from a reduction to the Longest Increasing Subsequence (LIS) problem. 
This reduction yields an efficient algorithm for generating sequences of moves $( s_1, s_2, \dots, s_d )$ of minimal size $d$, which we call \emph{schedules}. 
However, not all minimal size schedules incur the same  cost 
(i.e., number of column operations).
We investigate the feasibility of choosing optimal cost schedules, and show that 
greedy-type approaches can lead to arbitrarily bad behavior. 
In light of these results, we give an alternative proxy-objective for cost minimization, provide bounds justifying its relevance to the original problem, and give an efficient $O(d \log d)$ algorithm for approximately solving this proxy minimization. 
A performance comparison with other reduction-based persistent homology computations is given; 
our move schedules are demonstrated to be an order of magnitude more efficient than existing approaches in applications such as flock simulations and 2D persistence computations with RIVET. 
%We conclude with a discussion of situations where our strategy is most applicable and future directions for research.

@@

\subsection{Outline} The remainder of the paper is organized as follows: we review and establish the notations we will use to describe simplicial complexes, persistent homology, and dynamic persistence in Section \ref{sec:background}. 
We also cover the reduction algorithm (designated here as \texttt{pHcol}), the vineyards algorithm, and the set of \emph{move}-related algorithms introduced in~\cite{busaryev2010tracking}, which serves as the starting point of this work. 
In Section \ref{sec:move_schedules} we introduce our move schedules  and provide efficient algorithms to make this alternative viable. 
In Section \ref{sec:results} we present applications of the proposed method. 
In particular, to the computation of crocker stacks from flock simulations, and to 2-dimensional persistence on the Mumford data set of image patches from natural images. 
In Section \ref{sec:conclusion} we conclude the paper by discussing other possible applications and future work. 
We also include an appendix which introduces certain topics in more detail, such as the vineyards algorithm itself, for readers who may be unfamiliar but would like a more in-depth understanding.

@@

The $p$-boundary $\partial_p(\sigma)$  of a $p$-simplex $\sigma\in K$ is the alternating sum of its oriented co-dimension 1 faces,
and the $p$-boundary of a $p$-chain is defined 

@@

If $\{K_i\}_{i\in [m]}$ is a filtration, then the inclusion maps  $K_i\subset K_{i+1}$   induce linear transformations 
%$f_p^{i,j}: H_p(K_i) \to H_p(K_j)$ 
at the level of homology:

@@

%The $p$-th persistent homology groups are the images of these transformations: $H_{p}^{i,j} = \mathrm{Im}\,f_p^{i,j}$. 

@@

Let $\mathbb{X}$ be a triangulable topological space. 
That is, so that there exists an abstract 
simplicial complex $K$ whose geometric realization  is homeomorphic to 
$\mathbb{X}$. 
Let  $f: \mathbb{X} \to \mathbb{R}$ be continuous and  
write $\mathbb{X}_a = f^{-1}(-\infty, a]$ to denote the sublevel sets of $\mathbb{X}$ defined by the value $a$. 
A \emph{homological critical value} of $f$ is any value $a \in \mathbb{R}$ such that the homology of the sublevel sets of $f$ changes at $a$, i.e.  if for some $p$ the inclusion-induced homomorphism  $H_p(\mathbb{X}_{a - \epsilon}) \to H_p(\mathbb{X}_{a+\epsilon})$ is not an isomorphism for  any small enough $\epsilon >0$. If there are only finitely many of these homological critical values, then $f$ is said to be \emph{tame}.

Consider a homotopy $F(x,\tau) : \mathbb{X} \times [0,1] \to \mathbb{R}$ and denote its ``snapshot'' at a given time-point $\tau$ by $f_\tau(x) = F(x,\tau)$.
The snapshot $f_0$ denotes the initial function at time $\tau = 0$ and $f_1$ denotes the function at the last time step. 
As $\tau$ varies in $[0,1]$, the points in $\mathrm{dgm}_p(f_\tau)$ trace  curves in $\mathbb{R}^3$, called \emph{vines} and which together form a \emph{vineyard}. 
The stability of persistence implies that these curves will be continuous if $F$ is continuous
and the $f_\tau$'s are tame \cite{cohen2007stability}.
The vineyard analogy acts as a guidepost for practitioners seeking to understand how subtle changes occurring to the topological structure over time.
A more detailed discussion of the vineyards approach is provided in section~\ref{sec:vineyards}.
%Let $\mathbb{M}$ be a smooth $d$-manifold and $f : \mathbb{M} \to \mathbb{R}$ a smooth function. 
%A critical point $x$ is called \emph{non-degenerate} if the Hessian matrix of second partial derivatives is non-singular, otherwise if the Hessian is zero $x$ is said to be \emph{degenerate}. A \emph{Morse function} is a smooth function that has only non-degenerate critical points with distinct critical values. 
%and two Morse functions $f_0, f_1: \mathbb{M} \to \mathbb{R}$, and further suppose we equip $f_0, f_1$ with with a smooth homotopy $F : \mathbb{M}: [0,1] \to \mathbb{R}$ such that $F(x, 0) = f_0(x)$ and $F(x, 1) = f_1(x)$ for $x \in \mathbb{M}$. We use $f_\tau$ to denote the function at $F(x, \tau)$, where $\tau \in [0, 1]$. 
\subsection{The Reduction Algorithm}\label{sec:reduction}
In this section we briefly recount the original reduction algorithm introduced in~\cite{zomorodian2005computing}, also sometimes called the \emph{standard} algorithm or more explicitly \texttt{pHcol}~\cite{de2011dualities}. 
The pseudocode is outlined in Algorithm~\ref{alg:reduce}. Without optimizations, like   clearing   or       implicit matrix reduction, the standard  algorithm is very inefficient. Nonetheless, it serves as the foundation of most implementations that compute persistent homology, and its invariants are necessary before introducing both vineyards in section~\ref{sec:vineyards} and our move schedules  in section~\ref{sec:move_schedules}.

@@

Using $\mathbb{Z}_2$ coefficients and homology in dimension 1 to simplify the presentation, the reduction proceeds as follows:

@@

\noindent The column $c$ in $R$ indicates that dimension $1$ homology is born, which in this filtration is never killed since the dimension of the complex is $1$. Similarly, the columns at $u, v, w$ in $R_0$ (not shown) are all zero, indicating three $0$-dimensional homology classes  are born, which are then killed by the pivot entries in columns $a$ and $b$ in $R_1$.

@@

It is clear from Algorithm~\ref{alg:reduce} that a loose upper bound for this reduction is $O(m^3)$, where $m$ is the number of simplices of the filtration; it turns out that this bound is in fact tight, see~\cite{morozov2005persistence} for more details. 

@@

Notice that homological critical events can only occur when the filtration order changes, though the converse is not true.
Thus, detecting  all such events reduces to computing matrix decompositions at a finite set of time points interleaved between changes in the filtration order. 
At the finest scale, these changes manifest as transpositions of adjacent simplices. 
As a result, a fixed set of rules for maintaining a valid $R = D V$ decomposition under adjacent transpositions  is enough to simulate persistence dynamically. 

@@

 The time complexity of vineyards is determined entirely by the complexity of performing a single adjacent transposition.
 Informally, since column operations are the largest complexity operations needed and each column can have potentially $m$ entries, the complexity of vineyards is $O(m)$ per transposition. 
 However, there are a number of subtleties involved in this statement. 
 The first  is that both the $V$ and $R$ matrices tend to be quite sparse, and thus column and row operations rarely require $m$ operations---in fact, as a rule of thumb, most transpositions require no column operations~\cite{edelsbrunner2000topological}. A second subtlety, due to the necessity of sparse matrix data structures, is that querying the non-zero status of any particular entry typically takes $O(\log m)$ time. This hidden cost may seem minor, however, the quadratic scaling of the number of transpositions vineyards typically requires makes this cost non-trivial.   
In fact, achieving $O(m)$ complexity per transposition requires a special sparse matrix representation that allows swapping any two rows and columns in $O(1)$---see the appendix for more details and discussion. Finally, note the constant factor associated with the stated complexity: inspection of the individual cases of the algorihtm from~\cite{cohen2006vines} shows that there are at most two $O(m)$ operations, on both $R$ and $V$, that may be needed for any single transposition.

@@

Though they are similar, move operations confer a few computational advantages compared to vineyards. Informally, a move operation has two convenient properties:
\\	
 \begin{enumerate}[leftmargin=2\parindent, align=left, labelsep=-5pt, topsep=0pt,itemsep=-0.25ex,parsep=1.2ex]
 	\item[M1:] Querying the non-zero status of entries in either $R$ or $V$ occurs once per move.
 	\item[M2:] $R = D V$ is \underline{not} guaranteed to be valid  during the movement of $\sigma_i $ to $ \sigma_j$.
 \end{enumerate} 
 \vspace*{1em}
First, consider property M1. Prior to applying any permutation $P$ to the decomposition, it is necessary to remove non-zero entries in $V$ which render $P^TVP$ non-upper triangular, to maintain invariant I2. 
If one uses the (augmented) sparse matrix representation described in~\cite{cohen2006vines}, checking whether these entries are non-zero takes $O(\log m)$ per column. Since moving a simplex from $i$ to $j$ using vineyards performs $\lvert i - j \rvert - 1$ transpositions, this query operation is required at most $O(\lvert i - j \rvert)$ times.
By comparison, if a row-oriented sparse matrix implementation is used, then a move operation can perform these non-zero entry checks in just one $O(m)$ pass, prior to performing any column operations.

Property M2 indicates move operations only guarantee the decomposition is valid after the operation completes, implying that the decomposition is not fully maintained during the execution of \textit{RestoreRight} and \textit{RestoreLeft} below. 
This is a significant difference compared to the vineyards algorithm, which was designed specifically to maintain a valid decomposition after each transposition. 
In this way, we interpret move operations as making a tradeoff in granularity: whereas a sequence of adjacent transpositions $(i, i{+}1), (i{+}1, i{+}2), \dots, (j{-}1, j)$ generates $\lvert i - j \rvert $ valid decompositions in vineyards, a move operation $\mathrm{Move}(i,j)$ generates only one.
Indeed, if one has a pair of filtrations $(K_0, K_1)$ each with $m$ simplices, and we assume that each simplex $\sigma_i$ switches its relative ordering with another simplex $\sigma_j$ at most once in some ordered schedule of transpositions, then the number of intermediate persistence diagrams is bounded above by $O(m^2)$ using transpositions, whereas with move operations this bound clearly is at most $O(m)$.  \\
